---
title: "Module 8: Exploring species-environment relationships"
draft: false
layout: "single"
type: "page"
disableNewSection: true
---

<!-- Dropdown -->
<select id="module-select" class="course-dropdown" onchange="if (this.value) window.location.href=this.value;">
  <option value="">Jump to a module...</option>
  <option value="/r_for_nervous_humans/spatial/module-1/">1. What is spatial data?</option>
  <option value="/r_for_nervous_humans/spatial/module-2/">2. Opening and exploring spatial data</option>
  <option value="/r_for_nervous_humans/spatial/module-3/">3. Coordinates and projections</option>
  <option value="/r_for_nervous_humans/spatial/module-4/">4. Points, polygons, and rasters</option>
  <option value="/r_for_nervous_humans/spatial/module-5/">5. Are patterns real?</option>
  <option value="/r_for_nervous_humans/spatial/module-6/">6. Working with environmental data</option>
  <option value="/r_for_nervous_humans/spatial/module-7/">7. Preparing occurrence data for modelling</option>
  <option value="/r_for_nervous_humans/spatial/module-8/">8. Exploring species-environment relationships</option>
  <option value="/r_for_nervous_humans/spatial/module-9/">9. Building a species distribution model</option>
  <option value="/r_for_nervous_humans/spatial/module-10/">10. Assessing and predicting your SDM</option>
</select>

```{r echo = FALSE, message = FALSE, warning = FALSE}
library(sf)
library(terra)

load('data/region.RData')
load('data/module7_occdata.RData')
env_stack <- terra::rast('data/env_stack.tif')
```

### Exploring species-environment relationships

For this module, you'll need the following packages:

- `sf`
- `terra`

In Modules 7 and 8 we cleaned the occurrence data and built a combined data frame of presence and background points with their environmental values. In Module 9 we'll fit a formal model. But before we do any of that, we're going to *look* at the data.

This step is sometimes skipped in the rush to get to the modelling, but it's one of the most scientifically valuable things you can do. Exploratory analysis helps you understand which variables are likely to matter, form predictions about what the model should find, and catch problems before they cause cryptic errors downstream. A model fitted without prior exploration is a model you don't fully understand.

#### Load your data

```{r}
load('data/region.RData')
load('data/module7_occdata.RData')
env_stack <- terra::rast('data/env_stack.tif')

# Reminder of what we're working with
nrow(howler_sdm)
table(howler_sdm$occ)   # 1 = presence, 0 = background
```

#### 8.1 Where do howler monkeys occur environmentally?

The key question for an SDM is: what environmental conditions do howler monkeys occupy, and how does that differ from what's available across the landscape? Before fitting a model, we want a visual answer to this.

A good place to start is **boxplots** — one per variable, comparing values at presence points against background points. For each variable, a systematic shift between the two distributions is a signal that the variable might matter for the model.

```{r}
vars_to_plot <- c('bio01', 'bio04', 'bio12', 'bio15', 'elev')

par(mfrow = c(2, 3), mar = c(4, 4, 2, 1))

for (v in vars_to_plot) {
  boxplot(howler_sdm[[v]] ~ howler_sdm$occ,
          names  = c("Background", "Presence"),
          col    = c("grey80", "tomato"),
          ylab   = v,
          main   = v,
          outline = FALSE)
}
par(mfrow = c(1, 1))
```

For each variable, ask:

- Is the median shifted between presence and background?
- Is the spread (box width) different?
- Is the difference ecologically interpretable — does it match what you know about howler monkey biology?

Howler monkeys in Costa Rica are predominantly lowland animals, typically associated with warm, humid forest below about 1,500 m (Fedigan & Jack 2001). You might therefore expect presences to cluster at lower elevations and higher temperatures than background points — but check whether your data agrees.

#### 8.2 Density plots: seeing the full distribution

Boxplots summarise distributions but hide their shape. **Density plots** show more detail, including whether a distribution is unimodal (one peak), bimodal (two peaks), or skewed. This matters for modelling: a species with a hump-shaped response to temperature (most common at intermediate values) needs a quadratic term in the model; a monotonic response (simply more common at higher temperatures) can be captured with a linear term.

```{r}
par(mfrow = c(2, 3), mar = c(4, 4, 2, 1))

for (v in vars_to_plot) {
  pres_vals <- howler_sdm[[v]][howler_sdm$occ == 1]
  bg_vals   <- howler_sdm[[v]][howler_sdm$occ == 0]

  d_pres <- density(pres_vals, na.rm = TRUE)
  d_bg   <- density(bg_vals,   na.rm = TRUE)

  plot(d_bg, col = "grey60", lwd = 2,
       main = v, xlab = v,
       ylim = c(0, max(d_pres$y, d_bg$y) * 1.15))
  lines(d_pres, col = "tomato", lwd = 2)
  legend("topright", legend = c("Background", "Presence"),
         col = c("grey60", "tomato"), lwd = 2, bty = "n", cex = 0.8)
}
par(mfrow = c(1, 1))
```

Variables where the presence density (red) is a narrow peak sitting inside a broader background distribution (grey) are potentially very useful predictors — the species is selecting a subset of available conditions. Variables where the two curves almost completely overlap probably won't add much.

#### 8.3 Environmental space: two variables at once

Species don't experience temperature and rainfall in isolation — they experience the combination. A scatterplot of two variables, with presence and background points coloured differently, lets us see whether the species occupies a specific corner of **environmental space** (also called **climate space** or **niche space**).

```{r}
# Change these to whatever looked most different in your boxplots
v1 <- 'bio01'
v2 <- 'bio12'

plot(
  howler_sdm[[v1]][howler_sdm$occ == 0],
  howler_sdm[[v2]][howler_sdm$occ == 0],
  col  = adjustcolor("grey60", alpha.f = 0.3),
  pch  = 19, cex = 0.5,
  xlab = paste(v1, "(×0.1 °C)"),
  ylab = paste(v2, "(mm)"),
  main = "Environmental space — howler monkey"
)
points(
  howler_sdm[[v1]][howler_sdm$occ == 1],
  howler_sdm[[v2]][howler_sdm$occ == 1],
  col = adjustcolor("tomato", alpha.f = 0.7),
  pch = 19, cex = 0.7
)
legend("topright", legend = c("Background", "Presence"),
       col = c("grey60", "tomato"), pch = 19, bty = "n")
```

If the red cluster occupies a distinct region of the plot — say, high temperature and high precipitation — while background points are spread more broadly, that's a strong signal that the model has something real to learn. This visualisation is conceptually related to Hutchinson's (1957) ecological niche concept: the species occupies a subset of the available environmental space.

#### 8.4 Correlation among predictors

We have 20 potential predictors (19 bioclim variables plus elevation). Many are highly correlated with each other — mean annual temperature and mean temperature of the warmest quarter, for instance, will track each other very closely. Including redundant predictors doesn't improve a model; it can make it unstable.

Let's visualise correlations with a heatmap.

```{r}
env_vars <- howler_sdm[, 4:ncol(howler_sdm)]
cor_mat  <- cor(env_vars, method = 'spearman', use = 'complete.obs')

# Hierarchical clustering to order variables by similarity
ord         <- hclust(as.dist(1 - abs(cor_mat)))$order
cor_ordered <- cor_mat[ord, ord]

image(
  1:ncol(cor_ordered), 1:nrow(cor_ordered), t(cor_ordered),
  col  = colorRampPalette(c("steelblue", "white", "tomato"))(100),
  xaxt = "n", yaxt = "n", xlab = "", ylab = "",
  main = "Spearman correlation between predictors"
)
axis(1, at = 1:ncol(cor_ordered), labels = colnames(cor_ordered), las = 2, cex.axis = 0.7)
axis(2, at = 1:nrow(cor_ordered), labels = rownames(cor_ordered), las = 2, cex.axis = 0.7)
```

Deep red = strong positive correlation; deep blue = strong negative correlation; white = near-zero correlation. Clusters of highly correlated variables (blocks of red) are largely measuring the same thing. We'll deal with this formally in Module 9 using an automated variable selection procedure.

#### 8.5 A rough importance ranking

Before we fit anything, we can get a numerical sense of which variables are most different between presences and background — a simple proxy for how useful they might be.

```{r}
separation <- sapply(names(env_vars), function(v) {
  pres_mean <- mean(howler_sdm[[v]][howler_sdm$occ == 1], na.rm = TRUE)
  bg_mean   <- mean(howler_sdm[[v]][howler_sdm$occ == 0], na.rm = TRUE)
  bg_sd     <- sd(howler_sdm[[v]][howler_sdm$occ == 0],   na.rm = TRUE)
  abs(pres_mean - bg_mean) / bg_sd
})

sort(separation, decreasing = TRUE)
```

This standardised separation metric (effect size, roughly) ranks variables by how much their mean differs between presences and background, relative to background variability. Variables near the top of this list are worth prioritising in the model. Keep note of the top three or four — we'll see how they compare to what the formal variable selection procedure identifies in Module 9.

#### Exercise

1. Add `bio07` (temperature annual range) to your density plot loop. Is there a clear difference between presences and background? Given that howler monkeys live in tropical forest, would you expect them to be more sensitive to temperature seasonality or to mean temperature?

2. Make a scatterplot of `bio04` (temperature seasonality) versus `bio15` (precipitation seasonality). Do howler monkey occurrences cluster in a particular corner of this space?

3. Based on the correlation heatmap, identify two variables with very high correlation (|r| > 0.8). Then identify two variables with very low correlation. Which of the second pair would you expect to be more useful in an SDM, based on the density plots?

#### References

Fedigan, L.M. & Jack, K. (2001). Neotropical primates in a regenerating Costa Rican dry forest: a comparison of howler and capuchin population patterns. *International Journal of Primatology*, 22, 689–713.

Hutchinson, G.E. (1957). Concluding remarks. *Cold Spring Harbor Symposia on Quantitative Biology*, 22, 415–427.

<div style="margin-top: 2rem; display: flex; justify-content: space-between;">
  <a href="/r_for_nervous_humans/spatial/module-7/"
     style="padding: 0.6rem 1.2rem;
            background-color: var(--theme-accent);
            color: var(--theme-fg);
            text-decoration: none;
            border-radius: 6px;">
    ← Previous
  </a>
  <a href="/r_for_nervous_humans/spatial/module-9/"
     style="padding: 0.6rem 1.2rem;
            background-color: var(--theme-accent);
            color: var(--theme-fg);
            text-decoration: none;
            border-radius: 6px;">
    Next →
  </a>
</div>
