---
title: "Module 4 — Descriptive Statistics"
draft: false
type: "page"
layout: "single"
---

<!-- Dropdown -->
<select id="module-select" class="course-dropdown" onchange="if (this.value) window.location.href=this.value;">
  <option value="">Jump to a module...</option>
  <option value="/r_for_nervous_humans/">Welcome</option>
  <option value="/r_for_nervous_humans/module-1/">1. Getting Started</option>
  <option value="/r_for_nervous_humans/module-2/">2. Data & Penguins</option>
  <option value="/r_for_nervous_humans/module-3/">3. Visualisation</option>
  <option value="/r_for_nervous_humans/module-5/">5. Are These Groups Different?</option>
  <option value="/r_for_nervous_humans/module-6/">6. Working with Categories</option>
  <option value="/r_for_nervous_humans/module-7/">7. When Data Gets Weird</option>
  <option value="/r_for_nervous_humans/module-8/">8. Relationships Between Variables</option>
  <option value="/r_for_nervous_humans/module-9/">9. Linear Regression</option>
</select>

## Welcome to module 4

Now that we can look at penguin data and visualise them, it’s time to *summarise* our data numerically.  
Descriptive statistics help us answer questions like:

* What’s a “typical” value?  
* How much do values vary (differ)?  
* How confident are we about the average value of our sample? Does it accurately reflect the population?

Think of this as learning the language of the data before asking bigger questions.

---

## Prepare your dataset

Notice that we're now using an additional package, `ggplot2`. This supports more sophisticated plots and makes it easier to construct high-quality figures. 

```{r setup, warning=FALSE, message=FALSE}
library(palmerpenguins)
library(ggplot2)

data("penguins")

# Remove only the missing values for columns we need
penguins_clean <- penguins[!is.na(penguins$bill_length_mm) & 
                            !is.na(penguins$species) &
                            !is.na(penguins$body_mass_g), ]

```

Here, the square brackets mean 'look inside this thing'. Square brackets support indexing with rows separated from columns by a comma. So:

```{r}
penguins_clean[1,]
```

...gives you row 1. Similarly, 

```{r}
penguins_clean[,1]
```

...gives you column one. And,

```{r}
penguins_clean[1,1]
```

...gives you the cell found at row 1, column 1. You can use more complicated arguments to extract multiple values using an array argument `:` for concurrent data or `c()` for non-concurrent data. If that sounds like complete gibberish:

```{r}
penguins_clean[,2:5]
penguins_clean[,c(2,5)]
```

The first argument returns all of columns 1 to 5. The second only returns columns 2 and 5. 

```{r eval = FALSE}
penguins_clean <- penguins[!is.na(penguins$bill_length_mm) & 
                            !is.na(penguins$species) &
                            !is.na(penguins$body_mass_g), ]
```

So here we're looking within `penguins` and excluding rows (using `!is.na()`) where there are NAs in the columns `bill_length_mm`, 'species', and 'body_mass_g`. We are *not* selecting so are including them all.

---

### Measures of central tendency

**Mean** — the “typical” value

```{r}
mean(penguins_clean$bill_length_mm)
```

The mean is the foundation for most statistical tests like t-tests and regression. It has mathematical properties that make differences and variation easy to compute. Later, we’ll build everything from this anchor.

**Median** — the middle value

```{r}
median(penguins_clean$bill_length_mm)
```

The median is the value in the middle of the sorted data. Unlike the mean, it’s robust to extreme values.

---

### Spread — how much the data varies

**Standard deviation (SD)**

```{r}
sd(penguins_clean$bill_length_mm)
```

The Standard Deviation:
* Shows the typical distance of a given data point from the mean (i.e., the spread of the data)
* Small SD = tightly clustered; large SD = spread out.

**Standard error (SE)**

```{r}
se <- sd(penguins_clean$bill_length_mm) / sqrt(length(penguins_clean$bill_length_mm))
se
```

The Standard Error is computed by dividing the SD by the square root of the number of measurements in the sample. It shows us how accurately we can estimate the true average of the whole population based on our sample.
* Smaller SE = more confidence in the mean (i.e., our sample mean is close(r) to the population mean)

**95% confidence interval (CI)**

```{r}
mean_val <- mean(penguins_clean$bill_length_mm)
ci_lower <- mean_val - 1.96 * se
ci_upper <- mean_val + 1.96 * se
c(ci_lower, ci_upper)
```

The CI gives a plausible range for the true population mean. We typically use the 95% threshold as this  corresponds with our statistical probability threshold of p = 0.05. It basically indicates that if we repeated the study many times, around 95% of all calculated intervals would contain the true mean.

---

### When to use SD, SE, CI

| Measure | What it tells you | When to use |
|---------|-----------------|-------------|
| SD      | Spread of raw data | Understanding variability |
| SE      | Precision of the mean | Comparing sample means |
| 95% CI  | Range of true mean | Reporting uncertainty |

### Visual comparison
The code below may seem daunting but we'll cover `ggplot` in more detail later. For now, just notice the way in which plots can be built in this system by adding elements and using consistent language.

```{r}
set.seed(123)
x <- rnorm(30, mean = 40, sd = 5)

mean_x <- mean(penguins_clean$body_mass_g)
sd_x <- sd(penguins_clean$body_mass_g)
se_x <- sd_x / sqrt(length(penguins_clean$body_mass_g))
ci_x <- c(mean_x - 1.96*se_x, mean_x + 1.96*se_x)


# Create a data frame for plotting whiskers
plot_df <- data.frame(
  measure = c("SD", "SE", "95% CI"),
  lower = c(mean_x - sd_x, mean_x - se_x, ci_x[1]),
  upper = c(mean_x + sd_x, mean_x + se_x, ci_x[2])
)

# Plot
ggplot(plot_df, aes(x = measure)) +
  geom_point(aes(y = mean_x), size = 3, color = "blue") +          # Mean as point
  geom_errorbar(aes(ymin = lower, ymax = upper), width = 0.2, color = "red") +
  labs(title = "Mean with SD, SE, and 95% CI Whiskers",
       y = "Body mass (g)", x = "Measure") +
  theme_minimal() +
  geom_hline(yintercept = mean_x, linetype = "dashed", color = "blue") +
    # Annotations
  annotate("text", x = 2.3, y = mean_x[1], 
           label = "SD = data spread \n around mean", color = "black", size = 3.5, hjust = 0.5) +
  annotate("text", x = 3.4, y = mean_x[1], 
           label = "SE = accuracy \n of mean \n estimate", color = "black", size = 3.5, hjust = 0.5) +
  annotate("text", x = 1.4, y = mean_x[1], 
           label = "95% CI = likely \n range for true mean", color = "black", size = 3.5, hjust = 0.5) +
  
  annotate("text", x = 0.5, y = mean_x + 50, label = "Mean", color = "blue", hjust = 0)
```

You'll notice that these estimates are for all penguins. This might be an issue...

### Summarising by species

Before we start comparing penguins’ measurements statistically, it helps to look at each group separately. Why? Because treating all penguins as one giant lump ignores the fact that there may be important differences between species.

Summarising by group lets us see patterns and differences visually and numerically, which is exactly what statisticians do before running formal tests.

Let’s calculate the mean bill length for each species:

```{r}
tapply(penguins_clean$bill_length_mm,
       penguins_clean$species,
       function(x) mean(x, na.rm = TRUE))
```

What’s happening here:
1. `tapply()` splits the data into groups — here, by species.
2. The anonymous function `function(x) mean(x, na.rm = TRUE)` calculates the mean of each group, ignoring missing values.
3. The result is a small summary table showing average bill length per species.

You can also summarise multiple statistics at once:

```{r}
aggregate(bill_length_mm ~ species, data = penguins_clean, 
          FUN = function(x) c(mean = mean(x), sd = sd(x), n = length(x)))
```

Now we get the mean and SD for bill length and sample size per species in one glance. This gives a preview of what later t-tests or ANOVAs will formally test.

Let's adjust that plot from earlier...

```{r, echo=FALSE,  message = FALSE, warning = FALSE}
library(dplyr)
library(tidyr)
library(ggplot2)

# Summarise stats per species
plot_df <- penguins_clean %>%
  group_by(species) %>%
  summarise(
    mean_val = mean(body_mass_g, na.rm = TRUE),
    sd_val   = sd(body_mass_g, na.rm = TRUE),
    se_val   = sd_val / sqrt(n()),
    ci_lower = mean_val - 1.96 * se_val,
    ci_upper = mean_val + 1.96 * se_val
  ) %>%
  # Prepare ymin/ymax and offsets for each measure BEFORE pivoting
  mutate(
    sd_ymin = mean_val - sd_val,
    sd_ymax = mean_val + sd_val,
    se_ymin = mean_val - se_val,
    se_ymax = mean_val + se_val,
    ci_ymin = ci_lower,
    ci_ymax = ci_upper,
    sd_x = as.numeric(factor(species)) - 0.2,
    se_x = as.numeric(factor(species)),
    ci_x = as.numeric(factor(species)) + 0.2
  )

# Convert to long format for plotting
whiskers <- tibble(
  species = rep(plot_df$species, 3),
  x = c(plot_df$sd_x, plot_df$se_x, plot_df$ci_x),
  ymin = c(plot_df$sd_ymin, plot_df$se_ymin, plot_df$ci_ymin),
  ymax = c(plot_df$sd_ymax, plot_df$se_ymax, plot_df$ci_ymax),
  type = rep(c("SD", "SE", "95% CI"), each = nrow(plot_df)),
  mean_val = rep(plot_df$mean_val, 3)
)

# Plot
ggplot(whiskers, aes(x = x, color = type)) +
  geom_errorbar(aes(ymin = ymin, ymax = ymax), width = 0.1) +
  geom_point(aes(y = mean_val), size = 3) +
  scale_x_continuous(breaks = 1:3, labels = unique(plot_df$species)) +
  scale_color_manual(values = c("SD" = "red", "SE" = "green", "95% CI" = "purple")) +
  geom_hline(yintercept = mean(plot_df$mean_val), linetype = "dashed", color = "black") +
  labs(
    title = "Body Mass per Species with SD, SE, and 95% CI",
    x = "Species",
    y = "Body Mass (g)",
    color = "Whisker Type"
  ) +
  theme_minimal()
```

### Intuition check:
* Compare the means across species: are they very different?
* Look at the SDs: which species’ measurements are more variable?
* Notice the sample size: smaller groups give less precise estimates (we’ll see this in SE and CI).

By summarising by species, we’re building a clear picture of the data landscape. Later, when we run statistical tests, we’ll already have a solid understanding of our data.

### Lessons learned
* Mean is central for later tests
* Median shows the middle value
* SD measures spread of data
* SE measures precision of the mean
* 95% CI gives a plausible range for the true mean
* Plotting these together helps visualize differences
* Always summarise groups separately before testing