---
title: "Module 3 — Visualisation"
draft: false
type: "page"
layout: "single"
---

<!-- Dropdown for modules -->
<select id="module-select" class="course-dropdown" onchange="if (this.value) window.location.href=this.value;">
  <option value="">Jump to a module...</option>
  <option value="/r_for_nervous_humans/">Welcome</option>
  <option value="/r_for_nervous_humans/module-1/">1. Getting Started</option>
  <option value="/r_for_nervous_humans/module-2/">2. Data & Penguins</option>
  <option value="/r_for_nervous_humans/module-4/">4. Descriptive Statistics</option>
  <option value="/r_for_nervous_humans/module-5/">5. Are These Groups Different?</option>
  <option value="/r_for_nervous_humans/module-6/">6. Working with Categories</option>
  <option value="/r_for_nervous_humans/module-7/">7. When Data Gets Weird</option>
  <option value="/r_for_nervous_humans/module-8/">8. Relationships Between Variables</option>
  <option value="/r_for_nervous_humans/module-9/">9. Linear Regression</option>
</select>

### Welcome to Module 3

So far, we’ve looked *at* data. Now we’re going to start *seeing* it.

Numbers on their own are hard to interpret. A column of 300 values doesn’t naturally tell you a story. But a simple plot often does. Visualisation is often the fastest way to understand what’s going on. Before you start running tests or calculating anything complicated, a good plot can tell you:

Before we calculate anything complicated, it’s usually worth asking:

* Are values spread out or tightly grouped?  
* Are groups different?  
* Do two variables move together?  

Plotting data helps you answer questions like these.

By the end of this module, you will be able to:

* Create scatterplots, histograms, and boxplots  
* Understand what each type of plot shows  
* Begin spotting patterns in data  

---

```{r setup, include=FALSE}
library(palmerpenguins)
data("penguins")
```

### Histograms, or, 'what does this variable look like?'

A histogram shows how values are distributed.

```{r}
hist(penguins$bill_length_mm,
     main = "Distribution of Bill Length",
     xlab = "Bill Length (mm)",
     col = "skyblue",
     border = "white")
```

What to look for:
* Where are most values?
* Is the distribution symmetrical or skewed (i.e., larger on one side than the other)?
* Are there any unusual values?

### Boxplots — comparing groups

Boxplots are useful when you want to compare groups.

```{r}
boxplot(bill_length_mm ~ species,
        data = penguins,
        main = "Bill Length by Species",
        ylab = "Bill Length (mm)",
        col = c("lightpink","lightgreen","lightblue"))
```

Each box represents one group (here, a species of penguin). Things to look for:

* The middle line (median): where is the centre of each group?
* The height of the box: how spread out are the values?
* Any points outside the lines: potential outliers

At this stage, don’t worry about whether differences are “significant”. Just ask whether groups look different.

### Scatterplots — relationships between variables

Scatterplots show the relationship between two variables. Each point is one observation (one penguin, in this case).

Here, we’re asking: “As one variable changes, what happens to the other?”

```{r}
plot(penguins$flipper_length_mm,
     penguins$body_mass_g,
     main = "Flipper Length vs Body Mass",
     xlab = "Flipper Length (mm)",
     ylab = "Body Mass (g)",
     pch = 19,
     col = "steelblue")
```

Look for:
* An upward trend → as one increases, so does the other
* A downward trend → one increases while the other decreases
* No clear pattern → the variables might not be related

You’re not calculating anything yet — just noticing patterns.

---

### Adding colour (optional but useful)

Sometimes patterns aren’t obvious until you add a bit more information. `species` is already a factor (a categorical variable), which is great for analysis. When we produce basic plot, however, R needs **numbers to map to colours**, not category names (this changes with more advanced plotting packages but more on that, later). We convert the factor into numbers using `as.numeric()`, which lets R assign a different colour to each group:

```{r}
cols <- as.numeric(penguins$species)

plot(penguins$flipper_length_mm,
     penguins$body_mass_g,
     col = cols,
     pch = 19,
     xlab = "Flipper Length (mm)",
     ylab = "Body Mass (g)",
     main = "Flipper Length vs Body Mass (by Species)")
```

Now the plot tells a richer story. Instead of one cloud of points, you may start to see clusters. Intelligent use of colour can reveal patterns in the data. 

### Why this matters

Plots are not just for presentation — they’re an important part of thinking about data. Plots help you:

* Understand your data before analysing it  
* Spot patterns and relationships  
* Catch problems early (like outliers or strange values)  

Before doing any statistical test, it’s good practice to Look at your data *before* you test it.

---

### Lessons learned
* Histograms show distributions
* Boxplots compare groups
* Scatterplots show relationships
*Visualisation helps you understand data before analysing it

