---
title: "Module 6: Working with environmental data"
draft: false
layout: "single"
type: "page"
disableNewSection: true
---

<!-- Dropdown -->
<select id="module-select" class="course-dropdown" onchange="if (this.value) window.location.href=this.value;">
  <option value="">Jump to a module...</option>
  <option value="/r_for_nervous_humans/spatial/module-1/">1. What is spatial data?</option>
  <option value="/r_for_nervous_humans/spatial/module-2/">2. Opening and exploring spatial data</option>
  <option value="/r_for_nervous_humans/spatial/module-3/">3. Coordinates and projections</option>
  <option value="/r_for_nervous_humans/spatial/module-4/">4. Points, polygons, and rasters</option>
  <option value="/r_for_nervous_humans/spatial/module-5/">5. Are patterns real?</option>
  <option value="/r_for_nervous_humans/spatial/module-6/">6. Working with environmental data</option>
  <option value="/r_for_nervous_humans/spatial/module-7/">7. Preparing occurrence data for modelling</option>
  <option value="/r_for_nervous_humans/spatial/module-8/">8. Exploring species-environment relationships</option>
  <option value="/r_for_nervous_humans/spatial/module-9/">9. Building a species distribution model</option>
  <option value="/r_for_nervous_humans/spatial/module-10/">10. Assessing and predicting your SDM</option>
    <option value="/r_for_nervous_humans/spatial/module-11/">11. MaxEnt — maximum entropy modelling</option>
  <option value="/r_for_nervous_humans/spatial/module-12/">12. Random forest and ensemble models</option>
</select>

```{r echo = FALSE, message = FALSE, warning = FALSE}
library(sf)
library(terra)
library(geodata)
library(rnaturalearth)
library(rnaturalearthdata)

world  <- ne_countries(scale = "medium", returnclass = "sf")
region <- world[world$name == "Costa Rica", ]
region_vect <- vect(region)
region_bbox <- st_bbox(region)

geodata_path(".")
elev  <- elevation_global(res = 10)
elev_cr <- crop(elev, region_vect)
elev_cr <- mask(elev_cr, region_vect)

clim <- worldclim_global(var = 'bio', res = 10, download = FALSE, path = 'data')
names(clim) <- c(paste0('bio0', 1:9), paste0('bio', 10:19))
clim_cr <- crop(clim, region_vect)
clim_cr <- mask(clim_cr, region_vect)
```

# Working with environmental data

For this module, you'll need the following packages:

- `sf`
- `terra`
- `geodata`
- `rnaturalearth`
- `rnaturalearthdata`

In Module 5 we asked whether the patterns in our howler monkey data reflect real ecology or sampling effort. To start answering that — and to eventually build a species distribution model — we need environmental data describing what conditions actually exist across the landscape.

This module covers downloading, inspecting, and preparing the environmental layers we'll use throughout the rest of the course. By the end, you'll have a multi-layer raster stack covering Costa Rica that's ready for analysis.

# Why environmental data?

The dominant theory in species distribution modelling is **Hutchinson's n-dimensional hypervolume** — the idea that a species can persist anywhere that its full set of environmental requirements are met (Hutchinson 1957). In practice this is simplified considerably, but the core logic holds: species are found where their environment suits them. To predict where a species might occur, we need to know both where it has been recorded *and* what the environment looks like everywhere else.

The two most commonly used environmental data sources in SDM work are elevation and climate — particularly the **WorldClim** bioclimatic variables, which we'll use here.

# Loading packages and defining the study region

```{r eval = FALSE}
library(sf)
library(terra)
library(geodata)
library(rnaturalearth)
library(rnaturalearthdata)
```

```{r}
world       <- ne_countries(scale = "medium", returnclass = "sf")
region      <- world[world$name == "Costa Rica", ]
region_vect <- vect(region)       # terra needs SpatVector, not sf
region_bbox <- st_bbox(region)    # bounding box for cropping
```

# Elevation

We already met elevation in Module 4. Let's download it properly and prepare it for use alongside climate data.

```{r}
geodata_path(".")

elev    <- elevation_global(res = 10)
elev_cr <- crop(elev, region_vect)
elev_cr <- mask(elev_cr, region_vect)

plot(elev_cr, main = "Elevation — Costa Rica (m)")
```

```{r}
# Quick summary of elevation values
summary(values(elev_cr), na.rm = TRUE)
```

Costa Rica ranges from sea level to over 3,800 m at Cerro Chirripó. That's a remarkable range for a country roughly the size of West Virginia, and it's a large part of why Costa Rica is so biodiverse — it compresses multiple elevational zones and climate regimes into a small area (Janzen 1983).

# WorldClim bioclimatic variables

**WorldClim** provides 19 bioclimatic variables derived from monthly temperature and precipitation data, interpolated from global weather station networks (Fick & Hijmans 2017). They were designed specifically for species distribution modelling and are now one of the most widely used environmental datasets in ecology. The 19 variables cover things like:

- **bio01**: Mean annual temperature
- **bio04**: Temperature seasonality (standard deviation × 100)
- **bio12**: Annual precipitation
- **bio15**: Precipitation seasonality (coefficient of variation)
- **bio17**: Precipitation of the driest quarter

A full description of all 19 variables is at [worldclim.org/data/bioclim.html](https://www.worldclim.org/data/bioclim.html).

Note that temperature variables are stored in °C × 10 — so a value of 245 means 24.5°C. This is a legacy convention from when integer storage saved meaningful disk space. Divide by 10 to get degrees Celsius.

```{r}
# Set download = TRUE the first time, then FALSE to use cached files
clim <- worldclim_global(var = 'bio', res = 10, download = FALSE, path = 'data')

# Rename to cleaner labels
names(clim) <- c(paste0('bio0', 1:9), paste0('bio', 10:19))

# Crop and mask to Costa Rica
clim_cr <- crop(clim, region_vect)
clim_cr <- mask(clim_cr, region_vect)

clim_cr
```

Let's plot a few variables to get a feel for the data.

```{r}
# Mean annual temperature (divide by 10 for °C)
plot(clim_cr[['bio01']] / 10,
     main = "Mean annual temperature (°C)")

# Annual precipitation (mm)
plot(clim_cr[['bio12']],
     main = "Annual precipitation (mm)")

# Precipitation seasonality
plot(clim_cr[['bio15']],
     main = "Precipitation seasonality")
```

Notice how different variables tell different stories about the landscape. Temperature broadly tracks elevation — cooler highlands, warmer coasts. Precipitation is patchier, reflecting the complex rain shadow effects of the central mountains, which create a drier Pacific slope and a wetter Caribbean side.

# Checking that layers align

Before stacking and combining layers, we need to confirm that they share the same resolution, extent, and CRS. Layers that don't align can't be combined without error.

```{r}
# Resolution
res(elev_cr)
res(clim_cr)

# CRS
crs(elev_cr, describe = TRUE)
crs(clim_cr, describe = TRUE)
```

The resolutions should both be approximately 0.167° (10 arc-minutes). If they differ, we need to resample. Here we resample elevation to exactly match the climate grid, which ensures perfect alignment.

```{r}
elev_resamp <- terra::resample(elev_cr, clim_cr, method = 'bilinear')

# Confirm match
res(elev_resamp)
res(clim_cr)
```

Now we can stack them.

```{r}
env_stack <- c(clim_cr, elev_resamp)
names(env_stack)[20] <- 'elev'

nlyr(env_stack)   # should be 20
names(env_stack)
```

# Extracting values at points

To check everything's working, let's extract environmental values at a few howler monkey locations.

```{r}
library(rgbif)

species <- occ_search(
  scientificName = "Alouatta palliata",
  country = "CR",
  limit = 2000,
  hasCoordinate = TRUE
)

howler_sf <- st_as_sf(
  species$data,
  coords = c("decimalLongitude", "decimalLatitude"),
  crs = 4326
)
```

```{r echo = FALSE, message = FALSE, warning = FALSE}
howler_vect <- vect(howler_sf)

# Extract at first 5 locations
terra::extract(env_stack, howler_vect[1:5, ])
```

Each row is an occurrence point; each column is an environmental variable at that location. This is the foundation of species distribution modelling — the numerical pairing of "where was the species?" with "what was the environment there?"

# Save your work

We'll use these objects in every subsequent module.

```{r}
terra::writeRaster(env_stack, 'data/env_stack.tif', overwrite = TRUE)
terra::writeRaster(clim_cr,   'data/clim_cr.tif',   overwrite = TRUE)
# Note: region_vect is a terra SpatVector and cannot be saved with save() —
# terra's external pointer objects become invalid when serialised.
# We save region (an sf object) and recreate region_vect with vect() as needed.
save(region, region_bbox, file = 'data/region.RData')
```

# Exercise

1. Plot `bio04` (temperature seasonality). Where in Costa Rica is temperature most seasonal? Does this make geographical sense given what you know about the country's climate?
2. Use `terra::extract()` to get the mean annual temperature (`bio01 / 10`) at all howler locations. What is the range of temperatures where howlers have been recorded? Does this align with what you'd expect for a tropical forest species?
3. Use `terra::global(clim_cr[['bio12']], fun = 'mean', na.rm = TRUE)` to calculate mean annual precipitation across Costa Rica. Then do the same for `bio15` (precipitation seasonality). Which part of the country do you think is the most challenging environment for a forest-dependent species?

# References

Fick, S.E. & Hijmans, R.J. (2017). WorldClim 2: new 1-km spatial resolution climate surfaces for global land areas. *International Journal of Climatology*, 37, 4302–4315.

Hutchinson, G.E. (1957). Concluding remarks. *Cold Spring Harbor Symposia on Quantitative Biology*, 22, 415–427.

Janzen, D.H. (1983). No park is an island: increase in interference from outside as park size decreases. *Oikos*, 41, 402–410.

<div style="margin-top: 2rem; display: flex; justify-content: space-between;">
  <a href="/r_for_nervous_humans/spatial/module-5/"
     style="padding: 0.6rem 1.2rem;
            background-color: var(--theme-accent);
            color: var(--theme-fg);
            text-decoration: none;
            border-radius: 6px;">
    ← Previous
  </a>
  <a href="/r_for_nervous_humans/spatial/module-7/"
     style="padding: 0.6rem 1.2rem;
            background-color: var(--theme-accent);
            color: var(--theme-fg);
            text-decoration: none;
            border-radius: 6px;">
    Next →
  </a>
</div>
