---
title: "Module 1 — Getting Started"
type: "page"
---
  
  
<!-- Dropdown for modules -->
  <select id="module-select" class="course-dropdown" onchange="if (this.value) window.location.href=this.value;">
  <option value="">Jump to a module...</option>
  <option value="/r_for_nervous_humans/">Welcome</option>
  <option value="/r_for_nervous_humans/module-2/">Data & Penguins</option>
  <option value="/r_for_nervous_humans/module-3/">Visualisation</option>
  <option value="/r_for_nervous_humans/module-4/">Descriptive Statistics</option>
  <option value="/r_for_nervous_humans/module-5/">Are These Groups Different?</option>
  <option value="/r_for_nervous_humans/module-6/">Working with Categories</option>
  <option value="/r_for_nervous_humans/module-7/">When Data Gets Weird</option>
  <option value="/r_for_nervous_humans/module-8/">Relationships Between Variables</option>
  <option value="/r_for_nervous_humans/module-9/">Linear Regression</option>
</select>
  
  
  
  
### Welcome to Module 1
  
Ah, R. Some people see it and panic slightly, like a squirrel in a room full of rocking chairs.  
Fear not. Here, we take things one small step at a time.  

By the end of this module, you will be able to:
  
- Open R (or Posit Cloud) and run code  
- Understand what a variable is 
- Do basic arithmetic and store the results  

---
  
### Your R environment
  
#### Option 1: Posit Cloud 
  
Run R in the cloud. I don't recommend this long-term but it's a decent option to get you up-and-running quickly.

1. Go to [https://posit.cloud](https://posit.cloud)  
2. Create a free account  
3. Start a new R project  
4. Copy the code from this course and watch the magic happen  

#### Option 2: R on your own computer

1. Install [R](https://cran.r-project.org/)  
2. Install [RStudio](https://posit.co/download/rstudio-desktop/)  
3. Open RStudio  
4. Copy the code and enjoy the results  

RStudio is your *Graphical User Interface* (GUI) to R. Think of it like this: R is the engine, RStudio is the - not engine bits of the car. The bits that allow you to actually use the engine to do something useful. I don't know, I don't get cars.

RStudio has four main panes:
  
1. **Source (top-left)** – where you write scripts    
2. **Console (bottom-left)** – where R executes code  
3. **Environment / History (top-right)** – shows variables and what you’ve done  
4. **Files / Plots / Packages (bottom-right)** – your tools, graphs, and data 

It's important to understand that you can only save code written in your Source pane. 
 
---

## Opening a script

1. Open RStudio  
2. Go to **File → New File → R Script**  
3. A new blank tab appears in the **Source pane**  
4. Type your code here 

Scripts are like notebooks — you can save them, edit them, and re-run lines anytime.

---

## Running code

- Place your **cursor anywhere on the line** you want to run  
- Press **Ctrl + Enter** (Windows/Linux) or **Cmd + Enter** (Mac)  
- R will execute that line and show the result in the **Console pane**  

Tip: You can select multiple lines and run them all at once. 

---

### First tiny spells (arithmetic)

In R, you can do simple math, like this:

```{r eval = FALSE}
2 + 3
5 * 10
10 / 2
```

Hit CTRL+Enter after each line. Observe what R tells you. R always tells you the answer; but does it remember it? 

---

#### Variables: little boxes for numbers

Variables let you store a value for later. Imagine tiny boxes labeled with names like `x` or `y` and your values are contained in those boxes.

Run this code and see what happens. Rmember to look in your Global Environment, which is where variables are stored.

```{r}
x <- 10
y <- 5
z <- x + y
z
```

Note the <- symbol. This is the assignment arrow. It allows you to store the thing on the right as the thing on the left. 

Explanation:
* x <- 10 puts the value 10 into box x
* y <- 5 puts the value 5 into box y
* z <- x + y adds them and stores in box z
* Typing z shows the result

Exercise: Change the numbers in `x` and `y`. What does `z` become?

---

### Lessons learned

* R executes code line by line
* <- is the assignment arrow, putting values in boxes
* Variables can be used in calculations
* Experiment — that’s the fastest way to learn


<div style="margin-top: 2rem; display: flex; justify-content: space-between;">
  <a href="/r_for_nervous_humans/" 
     style="padding: 0.6rem 1.2rem; 
            background-color: var(--theme-accent); 
            color: var(--theme-fg); 
            text-decoration: none; 
            border-radius: 6px;">
    ← Previous
  </a>

  <a href="/r_for_nervous_humans/module-2/" 
     style="padding: 0.6rem 1.2rem; 
            background-color: var(--theme-accent); 
            color: var(--theme-fg); 
            text-decoration: none; 
            border-radius: 6px;">
    Next →
  </a>
</div>